---
title: "Losses Estimates from the Extreme Weather Condition in US"
author: "Cy"
date: "10/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Abstract
This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including the estimates of any fatalities, injuries, and property damage.    
This analysis is divided into two parts. Part 1 analyzes the fatalities and injuries caused by extreme weather conditions. Top 10 of event types are extracted and presented in graph.   Meanwhile, part 2 speaks about the Top 10 economic consequences in terms of property and agricultural crops damage.


## Data Processing
Download the data from website and read into "raw"
```{r reading, cache=TRUE}
if(!file.exists("storm.csv.bz2"))
        {
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                destfile = "storm.csv.bz2")
        }

raw<-read.csv(bzfile("storm.csv.bz2"))
```

Subsetting useful columns
```{r useful}
raweco<-raw[ ,c(8,23:28)]

```

Replacing all typos and similar event type into one category. ie: Whirlwind & WHIRLWIND
depending on 2.1.1 event table
```{r typos}
j<-list("WHIRLWIND","WATERSPOUT","VOLCANIC","FLOOD","THUNDERSTORM","SNOW")

for (i in 1:length(j)){
                raweco$EVTYPE[grepl(j[[i]],raweco$EVTYPE,ignore.case = TRUE)]<-j[[i]]
                }

raweco$EVTYPE<-factor(raweco$EVTYPE)

```


COnverting the prefix into value
```{r convert}
library(dplyr)
raweco$PROPDMGEXP<-recode(raweco$PROPDMGEXP,"K"=1000,"M"=1000000,"B"=1000000000,.default = 1)
raweco$CROPDMGEXP<-recode(raweco$CROPDMGEXP,"K"=1000,"M"=1000000,"B"=1000000000,.default = 1)
raweco$prop<-raweco$PROPDMG*raweco$PROPDMGEXP
raweco$crop<-raweco$CROPDMG*raweco$CROPDMGEX

```

## Results
Top 10 most harmful events
``` {r harming}
harm<-raweco%>%group_by(EVTYPE)%>%summarise('FATALITIES'=sum(FATALITIES),'INJURIES'=sum(INJURIES))%>% arrange(desc(FATALITIES+INJURIES))
                
print(harm[1:10,])

```

Extract top 10 and create plot

```{r plots}
library(ggplot2)
harm$EVTYPE<-factor(harm$EVTYPE)
harm10<-harm[1:10,]
p1<-ggplot(harm10)+geom_point(aes(EVTYPE,FATALITIES),color='red',size=2)+geom_point(aes(EVTYPE,INJURIES),color='blue',size=2)+theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+labs(y="Fatalities and Injuries")+labs(x="Type of Event")
print(p1)
```


  
  Top 10 most economical harmful events
``` {r harming2}
econ<-raweco%>% group_by(EVTYPE)%>% summarise('prop'=sum(prop),'crop'=sum(crop))%>% arrange(desc(prop+crop))
                
print(econ[1:10,])

```

Extract top 10 and create plot

```{r plot1}

econ$EVTYPE<-factor(econ$EVTYPE)
econ10<-econ[1:10,]
p2<-ggplot(econ10)+geom_point(aes(EVTYPE,prop),color='red',size=2)+geom_point(aes(EVTYPE,crop),color='blue',size=2)+theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+labs(y="Property and Crops Damage")+labs(x="Type of Event")
print(p2)
```
